Checkerboards And Missing Species Combinations: Are Ecological Communities

### 2016-01-28 ManuscriptMarkDown.Rmd
Added new analysis figures

---
title: 'Checkerboards And Missing Species Combinations: Are Ecological Communities
  Assembled By Chance?'
author: "Nicholas J. Gotelli"
date: "24 January 2016"
output:
  html_document:
    highlight: tango
    keep_md: yes
    theme: cerulean
  pdf_document: default
  word_document: default
---

@@

For example, here is a matrix illustrating the occurrence of the 27 species of finches (Subfamily Fringillidae) on the 25 major islands of the Greater and Lesser Antilles. Most of these species feed on seeds, which might be a shared, limiting resource that prevents all species from living together on the same island.

### Figure 1 here

```{r}
library(EcoSimR)
library(ggplot2)
library(reshape2)
set.seed(2016)
m <- melt(dataWiFinches)
m <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),m$value))

p <- ggplot( m, aes(x=V2, y=V1, fill=V3))
p + geom_tile()

Figure 1. A binary presence-absence matrix. Each row represents a different species of island bird (finches in the Subfamily Fringillidae), and each column represents a different island. Ecologists have often used these kinds of data matrices to infer whether species interactions are present.

In 1975, the ecologist Jared Diamond (best known for his 1999 Pulitzer-Prize-winning book *Guns, Germs, and Steel: The Fates of Human Societies*), asserted that these kinds of matrices contained the statistical signature of strong species interactions, which he boldly outlined as a series of community "assembly rules".

In one of his assembly rules, Diamond proposed that some pairs of species were such strong competitors that they never occured together on the same island, forming a "checkerboard distribution". For example, from the matrix above, we notice that the species in rows 3 and 6 (x and y) form a checkerboard pair and never co-occur:

### Figure 2 here

If we treat each island as a statistical replicate, these data can be analyzed as a 2 x 2 contingency table. Fisher's exact test can be used to ask whether the occurrence of species X is independent of that of species Y.

However, Diamond was not interested so much in the pattern for a single species in the assemblage. Rather, he argued that these checkerboard pairs could frequently be found in a community, and that they indicated competition for limited resources was important. For the West Indies finch matrix, there are 91 pairs of species that form perfect checkerboards.

@@

Is 91 checkerboards more than we would expect by chance? With 27 species, there are x unique pairs that can be formed, so perhaps finding 91/x pairs as checkerboards is not so extreme. Similarly, with 27 species, there are 2^27^ possible species combinations that could be formed. But with a sample of only 25 islands, the maximum possible value is only 25, which would happen if every island had a unique mixture of species. In this matrix, there are only 23 combinations that are observed. Is 23 combinations on 25 islands sampled unusually small, given that x combinations are theoretically possible?

In a provocative response to Diamond (1975), Connor and Simberloff (1979) introduced a Monte Carlo analysis (usually referred to as a "null model analysis" in this literature) to try and answer that question.

Connor and Simberloff asked, What would this matrix look like if each species colonized the different islands independently and species interactions were not important? Answering that question precisely would require a detailed ecological model with species-specific parameters for colonization, extinction, and persistence. There are many such models in the theoretical ecology literature, but getting the data to estimate all of those parameters is another story.

@@

Following standard Monte Carlo procedure, Connor and Simberloff randomized the observed presence-absence matrix, and calculated the number of checkerboard pairs and the number of species combinations that were found in this null matrix. Repeating this process 1000 times yields a distribution of the number of checkerboard pairs for this null hypothesis. Using a standard frequentist approach, we can then compare with the observed data and calculate how extreme or improbable the data are, given this null hypothesis as $(p(\text{Observed number of checkerboards} | H_0))$.

Exactly the same procedure could be used to evaluate the number of forbidden species combinations that are found in the matrix.

@@

The simplest initial approach is a kind of "isotropic" null hypothesis in which we simply reshuffle with equal probability all of the matrix elements. This null model preserves the dimensions of the matrix (the number of rows and columsn) and it preserves, the "fill" of the matrix, that is the proportion of presences (0.xx for the West Indices finch matrix). But there are no other constraints on the occurrence of species on islands.

@@

### Figure 3 here

@@

For these two models, the results are.....

>>>>

---
title: 'Checkerboards And Missing Species Combinations: Are Ecological Communities
  Assembled By Chance?'
author: "Nicholas J. Gotelli"
date: "24 January 2016"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
  html_document:
    fig_caption: yes
    highlight: tango
    theme: cerulean
---

@@

For example, here is a matrix illustrating the occurrence of the 17 species of finches (Subfamily Fringillidae) on the 19 major islands of the Greater and Lesser Antilles. Most of these species feed on seeds, which might be a shared, limiting resource that prevents all species from living together on the same island.
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 1. A binary presence-absence matrix. Each row represents a different species of island bird (finches in the Subfamily Fringillidae), and each column represents a different island. Ecologists have often used these kinds of data matrices to infer whether species interactions are present."}
library(EcoSimR)
library(ggplot2)
library(grid)
library(reshape2)
library(knitr)
set.seed(2016)
m <- melt(dataWiFinches)
m <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),m$value))

p <- ggplot(m, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none") + xlab("Islands") + ylab("Species")
print(p) 
```

Is the pattern of presences and absences in this matrix random? If so, does it reflect species interactions? These kinds of matrices have formed a kind of Rorschach Test for community ecologists. In 1975, the ecologist Jared Diamond (best known for his 1999 Pulitzer-Prize-winning book *Guns, Germs, and Steel: The Fates of Human Societies*), asserted that these kinds of matrices contained the statistical signature of strong species interactions, which he boldly outlined as a series of community "assembly rules".

One of Diamond's assembly rules was that some pairs of species were such strong competitors that they never occured together on the same island, forming a "checkerboard distribution". For example, from the matrix above, we notice that the species in rows 8 and 10 (*Loxigilla noxis* and *Tiara olivacea*) form a checkerboard and never co-occur:

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.cap=" Figure 2. A perfect bioegeographic checkerboard: each island supports one species, or the other, but never both."}
m <- melt(dataWiFinches[c(8,10),])
m <- as.data.frame(cbind(rep(1:2,19),rep(1:19,each=2),m$value))

p <- ggplot( m, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none")  + xlab("Islands") + ylab("Species")  + scale_y_continuous(breaks=c(1,2))
print(p)
```

The data can be organized into a two-way contingency table:

```{r echo=FALSE, warning=FALSE, message=FALSE}
Mychecker <- dataWiFinches[c(8,10),-1]
Mychecker <- as.matrix(Mychecker)
MyCon <- matrix(0,nrow=2,ncol=2, dimnames = list(c("Species A Present","Species A Absent"),c("Species B Present","Species B Absent")))
MyCon[1,1] <- sum(Mychecker[1,]==1 & Mychecker[2,]==1)
MyCon[1,2] <- sum(Mychecker[1,]==1 & Mychecker[2,]==0)
MyCon[2,1] <- sum(Mychecker[1,]==0 & Mychecker[2,]==1)
MyCon[2,2] <- sum(Mychecker[1,]==0 & Mychecker[2,]==0)
kable(MyCon, padding=0)
```

and a Fisher's exact test of the null hypothesis that the presence of Species A is independent of the presence of Species B yields p = `r fisher.test(MyCon)$p.value`.

However, Diamond did not emphasize the pattern for a single species in the assemblage. Rather, he argued that these checkerboard pairs could frequently be found in a community, and that the frequent occurrence of such pairs indicated competition for limited resources was generally important in organizing communities. For the West Indies finch matrix, there are 91 pairs of species that form perfect checkerboards.

@@

Is 91 checkerboards more than we would expect by chance? With 17 species, there are `r 17*16/2` unique pairs that can be formed, so perhaps finding $91/136$ = `r 91/136` checkerboards pairs of species is not so extreme. Similarly, with 17 species, there are $2^{17}$ = `r 2^17` possible species combinations that could be formed. But with a sample of only 19 islands, the maximum possible value is only 19, which would happen if every island had a unique mixture of species. The observed number was 10. Is 10 combinations represented in a sample of 19 islands sampled unusually small, given that $2^{17}$ combinations are theoretically possible?

In a provocative response to Diamond (1975), Connor and Simberloff (1979) introduced a Monte Carlo analysis (usually referred to  by ecologists as a "null model analysis") to address that question.

Connor and Simberloff asked: what would this matrix look like if each species colonized the different islands independently and species interactions were not important? Answering that question precisely would require a detailed ecological model with species-specific parameters for colonization, extinction, and persistence. There are many such models in the theoretical ecology literature, but getting the data to estimate all of those parameters is another story.

@@

Following standard Monte Carlo procedure, Connor and Simberloff randomized the observed presence-absence matrix, and calculated the number of checkerboard pairs and the number of species combinations that were found in this null matrix. Repeating this process 1000 times yields a distribution of the number of checkerboard pairs for this null hypothesis. Using a standard frequentist approach, we can then compare with the observed data and calculate how extreme or improbable the data are, given this null hypothesis as $(p(\text{Observed number of checkerboards} | H_0))$. The same procedure could be used to evaluate the number of species combinations that are found in the matrix.

@@

The simplest initial approach is a kind of "isotropic" null hypothesis in which we simply reshuffle with equal probability all of the matrix elements. This null model preserves the dimensions of the matrix (the number of rows and columsn) and it preserves, the "fill" of the matrix, that is the proportion of presences (0.xx for the West Indices finch matrix). But there are no other constraints on the occurrence of species on islands. Here is one such matrix, created by randomizing the occurrences in the original finch matrix:

```{r echo=FALSE, warning=FALSE, message=FALSE}
MyModel <- cooc_null_model(speciesData=dataWiFinches,metric="species_combo", algo="sim1",   suppressProg=TRUE)
m.combo <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),as.vector(MyModel$Randomized.Data)))

Mat1 <- ggplot(m.combo, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none")  + xlab("Islands") + ylab("Species")

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.cap=" Figure 3. An unconstrained randomization of the data in Figure 1."}
MyModel2 <- cooc_null_model(speciesData=dataWiFinches,metric="checker", algo="sim1",   suppressProg=TRUE)
m.checker <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),as.vector(MyModel2$Randomized.Data)))

Mat2 <- ggplot(m.checker, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none")  + xlab("Islands") + ylab("Species")

print(Mat2)
```

@@

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 4. Histogram of the number of species combinations and the number of checkerboard species pairs for the data in Figure 1. The blue vertical line indicates the observed data. By this test, the West Indies support more checkerboard species pairs but fewer species combinations than expected by chance."}
MyDat <- as.data.frame(MyModel$Sim)
binsize <- diff(range(MyModel$Sim))/4
Histo1 <- ggplot(MyDat, aes(x=MyModel$Sim)) + geom_histogram(binwidth=binsize,fill="goldenrod", colour="black",origin=5) + xlab("Number Of Species Combinations") + annotate("segment", x=10,xend=10,y=0,yend=630, colour="steelblue",size=1)

MyDat2 <- as.data.frame(MyModel2$Sim)
binsize <- diff(range(MyModel2$Sim))/15
Histo2 <- ggplot(MyDat, aes(x=MyModel2$Sim)) + geom_histogram(binwidth=binsize,fill="goldenrod", colour="black",origin=5) + xlab("Number Of Checkerboard Pairs") + annotate("segment", x=91,xend=91,y=0,yend=200, colour="steelblue",size=1)
multiplot(Histo1, Histo2,cols=2)
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE}
MyModel3 <- cooc_null_model(speciesData=dataWiFinches,metric="checker", algo="sim3",   suppressProg=TRUE)
m.checker <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),as.vector(MyModel3$Randomized.Data)))

Mat3 <- ggplot(m.checker, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none")  + xlab("Islands") + ylab("Species")

MyDat3 <- as.data.frame(MyModel3$Sim)
binsize <- diff(range(MyModel3$Sim))/15
Histo3 <- ggplot(MyDat3, aes(x=MyModel3$Sim)) + geom_histogram(binwidth=binsize,fill="goldenrod", colour="black") + xlab("Checkerboards") + annotate("segment", x=91,xend=91,y=0,yend=325, colour="steelblue",size=1)

MyModel4 <- cooc_null_model(speciesData=dataWiFinches,metric="checker", algo="sim2",   suppressProg=TRUE)
m.checker <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),as.vector(MyModel4$Randomized.Data)))

Mat4 <- ggplot(m.checker, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none")  + xlab("Islands") + ylab("Species")

MyDat4 <- as.data.frame(MyModel4$Sim)
binsize <- diff(range(MyModel4$Sim))/15
Histo4 <- ggplot(MyDat4, aes(x=MyModel4$Sim)) + geom_histogram(binwidth=binsize,fill="goldenrod", colour="black") + xlab("Checkerboards") + annotate("segment", x=91,xend=91,y=0,yend=220, colour="steelblue",size=1)

MyModel5 <- cooc_null_model(speciesData=dataWiFinches,metric="species_combo", algo="sim3",   suppressProg=TRUE)
m.checker <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),as.vector(MyModel5$Randomized.Data)))

Mat5 <- ggplot(m.checker, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none")  + xlab("Islands") + ylab("Species")

MyDat5 <- as.data.frame(MyModel5$Sim)
binsize <- diff(range(MyModel5$Sim))/4
Histo5 <- ggplot(MyDat5, aes(x=MyModel5$Sim)) + geom_histogram(binwidth=binsize,fill="goldenrod", colour="black") + xlab("Combinations") + annotate("segment", x=10,xend=10,y=0,yend=780, colour="steelblue",size=1)

MyModel6 <- cooc_null_model(speciesData=dataWiFinches,metric="species_combo", algo="sim2",   suppressProg=TRUE)
m.checker <- as.data.frame(cbind(rep(1:17,19),rep(1:19,each=17),as.vector(MyModel6$Randomized.Data)))

Mat6 <- ggplot(m.checker, aes(x=V2, y=V1, fill=V3)) + geom_tile() + theme(legend.position="none")  + xlab("Islands") + ylab("Species")

MyDat6 <- as.data.frame(MyModel6$Sim)
binsize <- diff(range(MyModel6$Sim))/8
Histo6 <- ggplot(MyDat6, aes(x=MyModel6$Sim)) + geom_histogram(binwidth=binsize,fill="goldenrod", colour="black") + xlab("Combinations") + annotate("segment", x=10,xend=10,y=0,yend=330, colour="steelblue",size=1)

multiplot(Mat3,Mat4,Histo5,Histo6,Histo3,Histo4,cols=3)
```
